---
title: "PM566 Midterm Project"
author: "Yumeng Gao"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---

### Prepare the library--> DELETE?
```{r}
library(dplyr)
library(tidyverse)
library(data.table)
library(R.utils)
library(lubridate)
library(leaflet)
library(webshot)
library(cowplot)
```






# Data Wrangling

### Download dataset from CDC's website and read it in.

DASH YRBSS - Alcohol and Other Drug Use (MS): https://chronicdata.cdc.gov/Youth-Risk-Behaviors/DASH-YRBSS-Alcohol-and-Other-Drug-Use-MS-/bdg5-rnzz

* This dataset was subtracted from DASH - Youth Risk Behavior Surveillance System (YRBSS): Middle School (https://chronicdata.cdc.gov/Youth-Risk-Behaviors/DASH-Youth-Risk-Behavior-Surveillance-System-YRBSS/k5bc-k3g8).

```{r}
if (!file.exists("YRBSS_MS.csv")) {
  download.file(
    url = "https://chronicdata.cdc.gov/api/views/bdg5-rnzz/rows.csv?accessType=DOWNLOAD", "YRBSS_MS.csv", method = "libcurl", timeout  = 60)
}
ms= data.table::fread("YRBSS_MS.csv")
```


# Exploratory Data Analysis
1. Formulate a question
2. Read in the data
3. Check the dimensions and headers and footers of the data
```{r}
dim(ms)
head(ms)
tail(ms)
str(ms)
```



4. Check the variable types in the data
5. Take a closer look at some/all of the variables
6. Validate with an external source
7. Conduct some summary statistics to answer the initial question - Make exploratory graphs















```{r}

```

```{r}

```













































































































































